---
title: "Code Cracker"
date: "2026-02-11"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data

```{r}
# Read the data directly from online (GitHub in this case)
crops <- read.csv("https://sselp.github.io/sdawr//Crop_recommendation.csv")

# Check the data
head(crops) # first 6 records
```

# Task example

- How many columns in the dataset?
- How many rows in the dataset?
- Is there any NA in the dataset? (Hint: function `anyNA`)

Rule: finish the questions and put the answers together.

```{r}
# Example answer
nc <- ncol(crops)
nr <- nrow(crops)
has_na <- anyNA(crops)
if (paste0(nc, nr, has_na) == "82200FALSE") print("CRACKED!")
```

# Task 1: diversity check

- How many unique crop types are in the label column? (Hint: function `unique`)
- What is the name of the crop that appears first alphabetically? (Hint: function `sort`)
- What is the name of the crop that appears last alphabetically? (Hint: set a parameter in `sort`)

```{r}
crop_types <- unique(crops[["label"]])
n_crops <- length(crop_types)
first_crop <- sort(crop_types)[1]
last_crop <- sort(crop_types, decreasing = TRUE)[1]
if (paste0(n_crops, first_crop, last_crop) == "22applewatermelon") print("CRACKED!")
```

`dplyr`

```{r}
n_crops <- crops %>% pull(label) %>% unique() %>% length()
first_crop <- crops %>% pull(label) %>% unique() %>% sort() %>% `[`(1)
last_crop <- crops %>% pull(label) %>% unique() %>% sort(decreasing = TRUE) %>% `[`(1)
```

# Task 2: The "High N" club

- What is the maximum N (Nitrogen) value for maize? (Hint: `max`, `==`, like `crops$label == "apple"` or `crops$P == 0`)
- How many rows in the dataset have an N value of exactly 0?
- Which crop label has the highest single N value in the whole dataset? (If more than one appears, pick the first one)

```{r}
max_n <- max(crops[crops$label == "maize", "N"])
zero_n <- sum(crops$N == 0)
crop_mn <- crops[crops$N == max(crops$N), "label"][1]
if (paste0(max_n, zero_n, crop_mn) == "10027cotton") print("CRACKED!")
```

`dplyr`

```{r}
max_n <- crops %>% filter(label == "maize") %>% select(N) %>% max()
zero_n <- crops %>% filter(N == 0) %>% nrow()
crop_mn <- crops %>% filter(N == max(N)) %>% select(label) %>% slice(1)
```

# Task 3: Soil chemistry

- What is the mean pH across all rows? (Round to 2 decimals). (Hint: function `mean` and `round`)
- How many rows have an ph greater than 7?
- Which crop has the lowest minimum ph value?

```{r}
ph_mean <- round(mean(crops$ph), 2)
rows_7 <- sum(crops$ph > 7)
crop_lph <- crops[crops$ph == min(crops$ph), "label"]
if (paste0(ph_mean, rows_7, crop_lph) == "6.47468mothbeans") print("CRACKED!")
```

`dplyr`

```{r}
ph_mean <- crops %>% pull(ph) %>% mean() %>% round(2)
rows_7 <- crops %>% filter(ph > 7) %>% nrow()
crop_lph <- crops %>% filter(ph == min(ph)) %>% select(label)
```

# Task 4: Tropical conditions

- How many rows have a temperature > 30 and humidity > 80? (Hint: `&` like `crops$label == "apple" & crops$N == 0`)
- How many crop types found in those "Tropical" conditions?
- What is the average rainfall for that specific subset? (Round to 0 decimals).

```{r}
crops_sub <- crops[crops$temperature > 30 & crops$humidity > 80, ]
n_rows <- nrow(crops_sub)
n_crops <- length(unique(crops_sub$label))
avg_rainfall <- round(mean(crops_sub$rainfall), 0)
if (paste0(n_rows, n_crops, avg_rainfall) == "1183119") print("CRACKED!")
```

`dplyr`

```{r}
n_rows <- crops %>% filter(temperature > 30 & humidity > 80) %>% nrow()
n_crops <- crops %>% filter(temperature > 30 & humidity > 80) %>% 
    pull(label) %>% unique() %>% length()
avg_rainfall <- crops %>% filter(temperature > 30 & humidity > 80) %>% 
    pull(rainfall) %>% mean() %>% round(0)
```

# Task 5: The "Ideal" Rice

- What is the average temperature for rice? (Round to 1 decimal).
- What is the max humidity for "rice"? (Round to 1 decimal). (Hint: `max`)
- What is the min pH for "rice"? (Round to 1 decimal). (Hint: `min`)

```{r}
avg_temp <- round(mean(crops[crops$label == "rice", "temperature"]), 1)
max_hmd <- round(max(crops[crops$label == "rice", "humidity"]), 1)
min_ph <- round(min(crops[crops$label == "rice", "ph"]), 1)
if (paste0(avg_temp, max_hmd, min_ph) == "23.7855") print("CRACKED!")
```

`dplyr`

```{r}
avg_temp <- crops %>% filter(label == "rice") %>% 
    pull(temperature) %>% mean() %>% round(1)
max_hmd <- crops %>% filter(label == "rice") %>% 
    pull(humidity) %>% max() %>% round(1)
min_ph <- crops %>% filter(label == "rice") %>% 
    pull(ph) %>% min() %>% round(1)
```

# Class 8
## Warm up

1. Identify all rows where pH greater than 8 and crop is "chickpea".
2. For only those rows, replace the current rainfall value with the maximum Nitrogen (N) value in this subset.
3. From this subset, randomly select 5 rows with seed of 10.
    - Compute: mean(rainfall) + mean(temperature) using those 5 rows.
    - Round the final result to 0.
    - Report the resulting value out loud once you finish.
    
```{r}
# 1. Identify the target rows
targets <- crops$ph > 8 & crops$label == "chickpea"

# 2. Extract the subset for calculation
chickpea_sub <- crops[targets, ]

# 3. Get the max N from the subset and replace rainfall
max_n_sub <- max(chickpea_sub$N)
chickpea_sub$rainfall <- max_n_sub

# 4. Randomly select 5 rows with seed 10
set.seed(10)
random_rows <- sample(1:nrow(chickpea_sub), 5)
final_subset <- chickpea_sub[random_rows, ]

# 5. Final math and rounding
raw_result <- mean(final_subset$rainfall) + mean(final_subset$temperature)
final_val <- round(raw_result, 0)
final_val
```

## Control structure
### If-else

```{r}
# 1. Select row No. 70
row70 <- crops[70, ]

# 2. Run the logic
if (row70$label == "rice") {
  # Add 2 to the temperature
  mod_temp <- row70$temperature + 2
  
  if (mod_temp > 25) {
    print("Vulnerable to climate change")
  } else {
    print("Resistant to climate change")
  }
  
} else {
  print("Not rice")
}
```

### switch

```{r}
set.seed(10)
rows <- crops[sample(1:nrow(crops), 1), ]

switch(rows$label,
    pigeonpeas = "Drought tolerant",
    kidneybeans = "Nutritious",
    muskmelon = "Juicy",
    stop("Invalid value")
)
```

### for

```{r}
for (i in 1:10) {
  if (crops$ph[i] < 6.0 & crops$N[i] > 80) {
    print(paste("Row", i, ": Action Required"))
  } else {
    print(paste("Row", i, ": Stable"))
  }
}
```

### While

```{r}
i <- 1
while (i <= 10) {
  if (crops$ph[i] < 6.0 & crops$N[i] > 80) {
    print(paste("Row", i, ": Action Required"))
  } else {
    print(paste("Row", i, ": Stable"))
  }
  i <- i + 1  # Crucial step!
}
```

### Repeat

```{r}
i <- 1
repeat {
  if (crops$ph[i] < 6.0 & crops$N[i] > 80) {
    print(paste("Row", i, ": Action Required"))
  } else {
    print(paste("Row", i, ": Stable"))
  }
  
  i <- i + 1
  
  if (i > 10) {
    break  # Stops the loop after row 50
  }
}
```

### `apply` practice on crops dataset

```{r}
row_sums <- apply(crops[, c("N", "P", "K")], 1, sum)

col_ranges <- lapply(crops[, c("temperature", "humidity", "rainfall")], range)
col_ranges <- do.call(rbind, col_ranges)
col_ranges

crop_rain <- tapply(crops$rainfall, crops$label, mean)

# use lapply
crop_rain <- do.call(c, lapply(unique(crops$label), function(x){
    mean(crops[crops$label == x, "rainfall"])
}))

scores <- mapply(function(n, p) n + (p * 2), crops$N, crops$P)

# use lapply
scores <- do.call(c, lapply(1:nrow(crops), function(x) {
    crops$N[x] + crops$P[x] * 2
}))
```
