---
title: "Code Cracker"
date: "2026-02-11"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data

```{r}
# Read the data directly from online (GitHub in this case)
crops <- read.csv("https://sselp.github.io/sdawr//Crop_recommendation.csv")

# Check the data
head(crops) # first 6 records
```

# Task example

- How many columns in the dataset?
- How many rows in the dataset?
- Is there any NA in the dataset? (Hint: function `anyNA`)

Rule: finish the questions and put the answers together.

```{r}
# Example answer
nc <- ncol(crops)
nr <- nrow(crops)
has_na <- anyNA(crops)
if (paste0(nc, nr, has_na) == "82200FALSE") print("CRACKED!")
```

# Task 1: diversity check

- How many unique crop types are in the label column? (Hint: function `unique`)
- What is the name of the crop that appears first alphabetically? (Hint: function `sort`)
- What is the name of the crop that appears last alphabetically? (Hint: set a parameter in `sort`)

```{r}
# Your code here, note that you should use the specified name for the objects

if (paste0(n_crops, first_crop, last_crop) == "22applewatermelon") print("CRACKED!")
```

# Task 2: The "High N" club

- What is the maximum N (Nitrogen) value for maize? (Hint: `max`, `==`, like `crops$label == "apple"` or `crops$P == 0`)
- How many rows in the dataset have an N value of exactly 0?
- Which crop label has the highest single N value in the whole dataset? (If more than one appears, pick the first one)

```{r}
# Your code here, note that you should use the specified name for the objects

if (paste0(max_n, zero_n, crop_mn) == "10027cotton") print("CRACKED!")
```

# Task 3: Soil chemistry

- What is the mean pH across all rows? (Round to 2 decimals). (Hint: function `mean` and `round`)
- How many rows have an ph greater than 7?
- Which crop has the lowest minimum ph value?

```{r}
# Your code here, note that you should use the specified name for the objects

if (paste0(ph_mean, rows_7, crop_lph) == "6.47468mothbeans") print("CRACKED!")
```

# Task 4: Tropical conditions

- How many rows have a temperature > 30 and humidity > 80? (Hint: `&` like `crops$label == "apple" & crops$N == 0`)
- How many crop types found in those "Tropical" conditions?
- What is the average rainfall for that specific subset? (Round to 0 decimals).

```{r}
# Your code here, note that you should use the specified name for the objects

if (paste0(n_rows, n_crops, avg_rainfall) == "11822119") print("CRACKED!")
```

# Task 5: The "Ideal" Rice

- What is the average temperature for rice? (Round to 1 decimal).
- What is the max humidity for "rice"? (Round to 1 decimal). (Hint: `max`)
- What is the min pH for "rice"? (Round to 1 decimal). (Hint: `min`)

```{r}
# Your code here, note that you should use the specified name for the objects

if (paste0(avg_temp, max_hmd, min_ph) == "23.7855") print("CRACKED!")
```


# Class 7
## Control structure
### If-else

```{r}

```

### switch

```{r}

```

### for

```{r}

```

### While

```{r}

```

### Repeat

```{r}

```

### `apply` practice on crops dataset

```{r}

```

